---
title: "Introduction to StatLens"
author: "Ian Alloway"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to StatLens}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
library(StatLens)
```

## Overview

**StatLens** is an R package that provides a focused, cohesive toolkit for
*exploratory data analysis* (EDA).  Rather than reimplementing general-purpose
statistics, StatLens offers carefully designed wrappers and visualisations that
help analysts quickly understand the structure and quality of a dataset.

The package is organised around two complementary object systems:

| System | Classes | Use case |
|--------|---------|----------|
| **S3** | `statlens_describe`, `statlens_outliers`, `statlens_missing`, `statlens_cor` | Lightweight result containers per function |
| **S4** | `DataProfile` | A validated, all-in-one profile of a full data frame |

---

## 1. Descriptive Statistics with `describe()`

`describe()` is a generic function that dispatches on the class of its input.

### 1a. Numeric vector

```{r describe-numeric}
x <- c(2.1, 3.4, 2.9, 5.0, 4.2, NA, 15.0, 1.8)
d <- describe(x)
print(d)
```

The returned object contains mean, sd, quartiles, IQR, skewness and kurtosis.
Use `summary()` to extract a tidy data frame:

```{r describe-summary}
summary(d)
```

### 1b. Data frame

When called on a data frame, `describe()` processes every numeric column:

```{r describe-df}
d_cars <- describe(mtcars)
print(d_cars)
```

The `plot()` method gives a quick bar chart of column means:

```{r describe-plot, fig.cap="Column means from describe(mtcars)"}
plot(d_cars, main = "mtcars — Column Means")
```

---

## 2. Outlier Detection with `detect_outliers()`

StatLens supports three detection strategies:

* **`"iqr"`** — Tukey's 1.5 × IQR fence (default)
* **`"zscore"`** — absolute Z-score threshold
* **`"both"`** — union of the above two

```{r outliers}
set.seed(42)
y <- c(rnorm(80, mean = 50, sd = 5), 85, 15, 90)

out <- detect_outliers(y, method = "iqr")
print(out)
```

```{r outliers-plot, fig.cap="Outliers highlighted in red"}
plot(out, main = "IQR Outlier Detection")
```

Using the Z-score method with a tighter threshold:

```{r outliers-z}
out_z <- detect_outliers(y, method = "zscore", z_threshold = 2.5)
print(out_z)
```

---

## 3. Missing Data Analysis with `missing_summary()`

```{r missing}
df_miss <- data.frame(
  age    = c(25, NA, 30, NA, 45, 22, NA, NA),
  income = c(50000, 62000, NA, 48000, 71000, NA, 55000, 60000),
  score  = c(88, 92, 75, NA, 81, 79, 95, NA),
  city   = c("A", "B", "A", "C", "B", NA, "A", "C")
)

ms <- missing_summary(df_miss, threshold = 0.20)
print(ms)
```

```{r missing-plot, fig.cap="Missing data by column (red bars exceed threshold)"}
plot(ms, main = "Missing Data Profile")
```

---

## 4. Correlation Analysis with `correlation_heatmap()`

```{r corr, fig.cap="Pearson correlation heatmap for mtcars"}
cm <- correlation_heatmap(mtcars[, 1:7])
```

Spearman correlations can be requested with `method = "spearman"`:

```{r corr-spearman, fig.cap="Spearman correlation heatmap"}
cm_s <- correlation_heatmap(mtcars[, 1:7], method = "spearman")
```

---

## 5. Moment Statistics: `skewness()` and `kurtosis()`

These functions implement the bias-corrected formulae used by common
statistical software:

```{r moments}
set.seed(7)
normal_data <- rnorm(500)
skewed_data <- rexp(500, rate = 0.5)

cat("Normal data  — skewness:", round(skewness(normal_data), 3),
    " kurtosis:", round(kurtosis(normal_data), 3), "\n")
cat("Skewed data  — skewness:", round(skewness(skewed_data), 3),
    " kurtosis:", round(kurtosis(skewed_data), 3), "\n")
```

---

## 6. Normalisation with `normalize()`

Four scaling strategies are available:

```{r normalize}
v <- c(10, 20, 30, 40, 50)
normalize(v, "minmax")   # [0, 1]
normalize(v, "zscore")   # zero mean, unit variance
normalize(v, "robust")   # median / IQR
normalize(v, "log")      # log(v + 1)
```

`normalize()` also works on entire data frames, scaling every numeric column
while leaving non-numeric columns unchanged:

```{r normalize-df}
cars_z <- normalize(mtcars, "zscore")
round(colMeans(cars_z), 6)  # all effectively 0
```

The scale parameters used are stored as an attribute for traceability:

```{r normalize-params}
str(attr(cars_z, "scale_params")[1:3])
```

---

## 7. Full Data Profiles with `profile_data()` (S4)

The `profile_data()` function constructs a `DataProfile` S4 object that bundles
everything above into a single, validated data structure.

```{r profile}
dp <- profile_data(iris, name = "Anderson's Iris Dataset")
show(dp)
```

### Detailed statistics table

```{r profile-summary}
profile_summary(dp)
```

### Diagnostic plots

```{r profile-plots, fig.cap="Column means from the DataProfile"}
profile_plot(dp, which = "means")
```

```{r profile-missing, fig.cap="Missing data rates"}
profile_plot(dp, which = "missing")
```

```{r profile-outliers, fig.cap="Outlier counts per column"}
profile_plot(dp, which = "outliers")
```

---

## 8. Design Principles

### S3 vs S4

StatLens deliberately uses **both** class systems:

* **S3** is used for per-function result objects (`statlens_describe` etc.)
  because it is lightweight, familiar, and easily extended without boilerplate.
* **S4** is used for `DataProfile` because the multi-slot structure benefits
  from formal validation (`validObject()`), explicit slot documentation, and
  the cleaner `setGeneric`/`setMethod` dispatch model.

### No heavy dependencies

StatLens imports only base and recommended packages (`stats`, `graphics`,
`grDevices`, `methods`, `utils`).  This keeps installation friction minimal
and makes the package suitable for use in restricted computing environments.

---

## Session Info

```{r session}
sessionInfo()
```
